Predicting Survival Probability of NASA Aircraft Engines

### 2025-04-15 index.qmd
Update 4/15 pt2

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald Test can
be used to determine which ones are statistically significant
individually and may have an impact in a multivariate model. Generating
a univariate CPH model for each covariate and using the Wald test with a
significance level of $\alpha = 0.05$ will allow for determination of if
a covariate is significant. Additional tests for models will involve
using a step-wise regression and comparing the results based on the
assumptions that are met and the concordance index. Any covariates that
are determined to be not significant to the survival probability can be
removed from the final CPH model.

@@

Forest plots can be used to visualize the effects of each covariate in
the model on the hazard ratio. The resulting plot contains each
covariate on the y axis, with the estimated effect on the hazard ratio
on the the x axis. The plot contains whiskers demonstrating a 95%
confidence interval for the effect. A positive effect indicates a
positive correlation with the hazard ratio and the higher the number the
more severe the effect is on the hazard ratio.

@@

The concordance index will produce a value between 0 and 1. A value of 1
means all pairs are correctly ordered and a value 0 being the
alternative with no pairs correctly ordered. Any values between 0 and 1
indicate how accurate the CPH model is with the data [@Therneau2017].

### Survival Prediction

@@

Based on the above test, none of the covariates are statistically
significant individually with all the data. As such, to determine the
best model to use for this analysis, a forward and backward step-wise
regression will be done to generate the models with the best Akaike
Information Criterion (AIC) and Bayesian Information Criterion (BIC)
values and compare the concordance values of those models with models
that have covariates that were statistically significant when used in a
CPH model with the training data. The model with the highest concordance
value, and therefore, the most accurate will be used for the analysis.

The table below provides the model number, the covariates used, the AIC
and BIC from the step-wise regression if applicable, and the concordance
index. The concordance index generates a value to determine which model
is the most accurate with the provided covariates. After determining the
most accurate model, the CPH assumptions will be tested.

@@

train_sig <- lapply( univ_formulas, function(x){coxph(x, data = train_grouped)})

>>>>

Depending on the study, there will be covariates that do not impact the
probability of survival. Statistical methods such as the Wald Test can
be used to determine which ones are statistically significant
individually and may have an impact in a multivariate model. Generating
a univariate CPH model for each covariate and using the Wald test with a
significance level of $\alpha = 0.05$ will allow for determination of if
a covariate is significant. Additional tests for models will involve
using a stepwise regression and comparing the results based on the
assumptions that are met and the concordance index. Any covariates that
are determined to be not significant to the survival probability can be
removed from the final CPH model.

@@

Forest plots can be used to visualize the effects of each covariate in
the model has on the hazard ratio. The resulting plot contains each
covariate on the y axis, with the estimated effect on the hazard ratio
on the the x axis. The plot contains whiskers demonstrating a 95%
confidence interval for the effect. A positive effect indicates a
positive correlation with the hazard ratio where the higher the number
is the more severe the effect on the hazard ratio.

@@

The concordance index will produce a value between 0 and 1. A value of 1
means all pairs are correctly ordered and a value of 0 being the
alternative with no pairs correctly ordered. Any values between 0 and 1
indicate how accurate the CPH model is with the data [@Therneau2017].

### Survival Probability

@@

Based on the above test, none of the covariates are statistically
significant individually with all the data. As such, to determine the
best model to use for this analysis, a forward and backward stepwise
regression will be done to generate the models with the best Akaike
Information Criterion (AIC) and Bayesian Information Criterion (BIC)
values and compare the concordance values of those models with models
that have covariates that were statistically significant when used in a
CPH model with the training data. The model with the highest concordance
value, and therefore, the most accurate will be used for the analysis.

The table below provides the model number, the covariates used, the AIC
and BIC from the stepwise regression if applicable, and the concordance
index. After determining the most accurate model, the CPH assumptions
will be tested.

@@

### 2025-04-15 slides.qmd
Update 4/15 pt2

```{r, echo = FALSE}
#generation of the model for sample plots
example_surv_object <- Surv(time = lung$time, event = lung$status)

example_coxmodel <- coxph(example_surv_object ~ meal.cal, data = lung)
```

@@

## Linearity Assumption

-   Tested using the Martingale residuals for each covariate using the
    equation Martingale Residuals = Observed Events - Expected Results
    [@nahhas2025]

-   If the plots are linear and appear to have a slope of zero, the
    assumption is not violated [@Amini2015]

```{r}
#| fig-align: center
example_martingale <- resid(example_coxmodel, type = "martingale")

plot(example_martingale, 
     main = "Martingale Residuals: Example Covariate",
     ylab = "Martingale Residuals for Example Covariate",
     xlab = "Value Range for Example Covariate")
abline(h = 0, col = "red")

```

## Proportional Hazards Assumption

-   The test is verified using the Schoenfeld partial residuals which
    are the difference between the value of the covariate and the
    expected value of the covariate at the time of failure [@klein2005]

::: panel-tabset
#### Graph

```{r}
example_phtest <- cox.zph(example_coxmodel)

plot(example_phtest[1],
     ylab = "Covariate Schoenfeld Residual",
     xlab = "Time")
```

#### Table

```{r}
multi <- coxph(example_surv_object ~ meal.cal + age, data = lung)
ph_tibble <- cox.zph(multi) %>%
  pluck("table") %>%
  as.data.frame() %>%
  tibble::rownames_to_column("variable") %>%
  as_tibble()
kable(ph_tibble, format = "markdown")
```
:::

## Kaplan-Meier Survival Curve

::: incremental
-   Provides visualization for the Kaplan-Meier estimator

-   Is considered to be well defined up until the largest observed study
    time $t_{max}$

-   Demonstrates the time where the event being modeled is expected to
    occur
:::

```{r}
#| fig-align: center
example_curve <- survfit(example_coxmodel)

ggsurvplot(example_curve,
           data = lung,
           title = "Example Survival Curve",
           xlab = "Time",
           ylab = "Survival Probability",
           legend = "none")
```

## Forest Plot

-   Provides visualization of the effects of each covariate on the
    hazard ratio

-   A positive effect indicates a positive correlation with the hazard
    ratio

```{r, warning=FALSE}
#| fig-align: center
forest <- coxph(example_surv_object ~ age + sex, data = lung)

plot_model(forest,
           dot.size = 1,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")

```

## Evaluation

## Prediction

## Data Structure, Exploration, and Visualization

@@

@@

## Adjusting the Model

>>>>

@@

## Visualization

The two common model visualizations are:

::: incremental
-   Kaplan-Meier Survival Curve

    -   Demonstrates the time where the event being modeled is expected
        to occur [@klein2005]

-   Forest Plots

    -   Visualizes the effects of each covariate in the model on the
        hazard ratio
:::

## Evaluation and Survival Probability

::: incremental
-   Model accuracy evaluation is done using the concordance index

    ::: incremental
    -   The concordance index measures the amount of agreement between
        two variables

    -   A value of 1 means all the pairs are correctly ordered while a
        value of 0 means no pairs are correctly ordered
    :::

-   Survival probability can be predicted at a specific time $t$

    ::: incremental
    -   If the probability is $\geq 50$%, it is assumed the event has
        not occurred

    -   If the probability is $< 50$%, it is assumed the event has
        occurred
    :::
:::

## Data Structure

The data selected for this project comes from a study on propagation
modeling that NASA completed, specifically focusing on the engine two
testing and training datasets [@saxena2008].

-   Each engine in the NASA data has an unknown amount of wear,
    manufacturing variation, and sensor noise

-   There are three operation setting fields and twenty-one sensor
    measurement fields

-   A column indicating status was added to both the testing and
    training datasets with 0 indicating the machine has not failed and 1
    indicating the machine has failed

-   The data was then merged together to make one grouped dataset

```{r}
testurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/test_FD002.txt'

test <- read.delim(testurl, header = TRUE, sep = '') %>%
  mutate(status = 0,
         id = id + 260)

trainurl <- 'https://raw.githubusercontent.com/JR-87/capstone4ds_reed/refs/heads/main/train_FD002.txt'

train <- read.delim(trainurl, header = TRUE, sep = '') %>%
  group_by(id) %>%
  mutate(status = if_else(row_number() == n(), 1, 0)) %>%
  ungroup()

data_all <- rbind(train, test)

train_grouped <- train %>% 
  group_by(id) %>%
  slice_max(time)

data_grouped <- data_all %>%
  group_by(id) %>%
  slice_max(time)

temp <- data_grouped %>% select(id, time, status, os1, os2, os3, sm1, sm2, sm3)

kable(head(temp, n = 5), format = "markdown", caption = 'NASA Aircraft Engine Data')
```

## Exploration

-   519 engines in the combined data

    -   260 engines in training data

    -   259 engines in testing data

```{r}
summarymetrics <- data.frame(c("Minimum", "Median", "Mean", "Standard Deviation", "Maximum"),
                             c(min(train_grouped$time), 
                               median(train_grouped$time),
                               mean(train_grouped$time),
                               sd(train_grouped$time),
                               max(train_grouped$time)))

colnames(summarymetrics) <- c("Metric", "Value")
summarymetrics$Value <- round(summarymetrics$Value, digits = 2)
kable(summarymetrics, format = "markdown", align = "c", caption = "Summary Metrics")
```

## Visualization

::: panel-tabset
#### Histogram/Boxplot

```{r}
#| fig-align: center
Histogram <- ggplot(data = train_grouped,
       aes(x = time)) +
  geom_histogram(binwidth = 5,
                 fill = "grey")+
  labs(x = "Survival Time (Iterations)",
       y = "Frequency")+
  theme_bw()

box <- ggplot(data = train_grouped,
                  aes(y = time))+
  geom_boxplot(fill = "gray",
               color = "black")+
  ylab("Time (Iterations)")+
  theme_bw()

grid.arrange(Histogram,box,ncol = 2, top = "Distribution of Engine Longevity in Training Data")
```

#### Survival Curve

```{r}
#| fig-align: center
t <- as.matrix(train_grouped[, "time"])
s <- as.matrix(train_grouped[, "status"])

km_fit <- survfit(Surv(t, s) ~ 1)

autoplot(km_fit) + theme_bw()
```
:::

@@

The table provides the model number, the covariates used, the AIC and
BIC from the stepwise regression if applicable, and the concordance
index.

```{r}
surv_object <- Surv(data_grouped$time, data_grouped$status)

all_cov <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

aic <- MASS::stepAIC(all_cov, trace = 0)
n <- nrow(data_grouped)
bic <- step(all_cov, direction = "both", k = log(n), trace = 0)

cox1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm16 + sm17 + sm18 + sm19 + sm20 + sm21)

cox2 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm5 + sm8 + sm9 + sm15 + sm16 + sm17 + sm18 + sm20)

cox3 <- coxph(data = data_grouped, surv_object ~ os3 + sm3 + sm4 + sm8 + sm9 + sm15 + sm18)

cox4 <- coxph(data = data_grouped, surv_object ~ os3 + sm13 + sm14 + sm19)

stepwise_results <- data.frame(c("model 1", "model 2", "model 3", "model 4"), 
                               c("All covariates", "os3, sm3 - sm5, sm8 - sm9, sm15 - sm18, sm20", "os3, sm3 - sm4, sm8-sm9, sm15, sm18", "os3, sm13 - sm14, sm19"),
                               c("2449.92", "2431.88", "N/A", "N/A"),
                               c("2547.71", "N/A", "2461.31", "N/A"),
                               c(cox1$concordance[6], cox2$concordance[6], cox3$concordance[6], cox4$concordance[6]))

colnames(stepwise_results) <- c("Model", "Covariates", "AIC", "BIC", "Concordance")

kable(stepwise_results, format = "markdown")
```

Model for the continuing analysis will have all covariates except sm16
and sm19.

## CPH Model Forest Plot

```{r}
#| fig-align: center
cox1.1 <- coxph(data = data_grouped, surv_object ~ os1 + os2 + os3 + sm1 + sm2 + sm3 + sm4 + sm5 + sm6 + sm7 + sm8 + sm9 + sm10 + sm11 + sm12 + sm13 + sm14 + sm15 + sm17 + sm18 + sm20 + sm21)

ci <- cox1.1$concordance[6]

plot_model(cox1.1,
           dot.size = 2,
           line.size = 1,
           colors = "red")+
  theme_bw()+
  labs(title = "Estimated Effects of Covariates on Hazard Ratios")+
  ylab(label = "Covariates")+
  xlab(label = "Estimated Effect on Hazard Ratio")
```

@@
